Distributional Trees for Circular Data

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2424 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: af56616cddd904fb55a07715460f4b2eb289ca1c

% Brief abstract of the paper:
\abstract{
%Circular response variables occur in a variety of applications, e.g., when
%geographical directions are considered or measurements are taken on a $24$-hour scale. 
A commonly chosen distribution for modeling circular data is the von Mises distribution.
As an alternative to estimating its distribution parameters by generalized additive models 
we propose the application of distributional trees which are capable of detecting also 
non-additive effects, select covariates and their possible interactions automatically and 
provide easily interpretable models allowing for a wide range of inference procedures. 
This approach is illustrated in an application to wind directions with a set of meteorological
covariates, all measured at the airport of Innsbruck, Austria.}

@@

%\subsection{Von Mises Distribution}
%The von Mises distribution is also known as the circular normal distribution 
%and its density function is given by
%\begin{equation}
%  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
%\end{equation}
%with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
%modified Bessel function of the first kind and order $0$.
%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
%to the variance $\sigma^2$ of the normal distribution. 
%See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.

@@

In the here discussed application the MOB algorithm is applied and the prespecified 
distribution family is the von Mises distribution which is also known as the circular 
normal distribution. Its density function is given by
\begin{equation}
  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
with location parameter $\mu$, dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
modified Bessel function of the first kind and order $0$.
%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
%to the variance $\sigma^2$ of the normal distribution. 
(See e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview.)

In each node of the tree the distribution parameters $\mu$ and $\kappa$ are 
%vector $\theta = (\mu,\kappa)$ is 
estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$. 
For the resulting estimators $\hat{\mu}$ and $\hat{\kappa}$ model scores are calculated by 
$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$
yielding a $n \times 2$-matrix for a data set of $n$ observations.
%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
In that way a value indicating the goodness of fit is provided for each parameter and observation
such that changes can be detected not only in the location parameter $\mu$ but also in the 
dispersion parameter $\kappa$.

@@

As an illustration of the tree based wind direction forecasts at Innsbruck
airport, a BIC based post-pruned regression tree of depth three is shown in
Fig.~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
valley (from east to west, nodes 4 and 5). Secondly, downslope winds blowing
across the Alpine crest along the Wipp Valley towards Innsbruck (from
south-east to north-west, nodes 7 and 8). Thirdly, down-valley winds blowing
in the direction of the valley mouth (from west to east, nodes 12, 14, and 15).
The final node 11 with a concentration parameter $\kappa$ of only $0.84$
captures the cases with rather low wind speeds which typically don't show a
clear wind direction or distinct wind regime. In general, winds typically blow
from areas of high pressure to areas of low pressure where strong pressure
gradients yield high wind speeds and a rather well-defined wind direction.
Thus, the splits in the lagged wind direction perform a clustering into the
different wind regimes and the additional splits in pressure gradients or wind
speed divide the predictions according to the expected consistency of the wind
direction. This can be seen in the final nodes with strong pressure gradients
or high wind speeds where most training cases are classified correctly and
therefore the concentration parameter as an indicator of the certainty of the
forecast is rather large.

\section{Discussion / Summary and Outlook}
Applying distributional trees on circular data provides easily interpretable models clearly
illustrating the impact of given covariates and their possible interactions on the distribution
of the dependent variable. Moreover, covariates are selected automatically such that no 
prespecification of the model is required which can be a challenging task, particularly for a 
meteorlogical quantity such as wind direction with high variation and many possible influencing 
factors.

This application shows that including covariates such as lagged observations of wind direction
as linear regressors provides valuable information. However, in order to properly account not 
only for the circularity of the response but also of regressors, additional adaptions of 
distributional trees are required. In particular, if the selected split variable is a circular 
variable this demands for finding a ``split interval'' instead of a split point.
This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
the linear regressor \textit{dd} (lagged wind direction) are required in order to represent its effect.

@@

Another possible extension of the here presented methodology is to build not just one tree 
model but an ensemble of trees to regularize and stabilize the model 
(see e.g. distributional forests, Schlosser et. al, 2018). In that way, it should not have a 
significant impact on the results whether circular covariates are included as linear regressors
yielding a larger number of splits or as circular regressors with ``intervall splits''.

>>>>

% Brief abstract of the paper:
\abstract{
%Circular response variables occur in a variety of applications, e.g., when
%geographical directions are considered or measurements are taken on a $24$-hour scale.
A commonly chosen distribution for modeling circular data is the von Mises distribution.
As an alternative to estimating its distribution parameters by generalized additive models 
we propose the application of distributional trees which are capable of detecting also 
non-additive effects, select covariates and their possible interactions automatically and 
provide easily interpretable models allowing for a wide range of inference procedures. 
This approach is illustrated by an application of distributional trees on hourly wind 
direction forecasts with a set of meteorological covariates, all measured at the airport 
of Innsbruck, Austria.
}

@@

%\subsection{Von Mises Distribution}
%The von Mises distribution is also known as the circular normal distribution 
%and its density function is given by
%\begin{equation}
%  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
%\end{equation}
%with a location parameter $\mu$, a conentration parameter $\kappa$, and $I_0(\kappa)$ being the 
%modified Bessel function of the first kind and order $0$.
%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
%to the variance $\sigma^2$ of the normal distribution. 
%See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.

@@

In this study distributional trees are applied to circular response data employing the MOB algorithm
and the von Mises distribution which is also known as the circular 
normal distribution. Its density function is given by
\begin{equation}
  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
with location parameter $\mu$, conentration parameter $\kappa$, and $I_0(\kappa)$ being the 
modified Bessel function of the first kind and order $0$
%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
%to the variance $\sigma^2$ of the normal distribution. 
(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).

In each node of the tree the distribution parameters $\mu$ and $\kappa$ are 
%vector $\theta = (\mu,\kappa)$ is 
estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$. 
For the resulting estimators $\hat{\mu}$ and $\hat{\kappa}$ model scores are calculated by  
%$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
%\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$
$s(y,\hat{\mu}, \hat{\kappa}) = (\partial_{\mu} l(y, \hat{\mu}, \hat{\kappa}),
\partial_{\kappa}l(y, \hat{\mu}, \hat{\kappa}))$
yielding a $n \times 2$-matrix for a data set of $n$ observations.
%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
In that way for each parameter a value indicating the goodness of fit with respect to one single 
observation is provided such that changes can be detected not only in the location parameter $\mu$ 
but also in the concentration parameter $\kappa$.

@@

As an illustration of the tree based wind direction forecasts at Innsbruck
airport, a BIC based post-pruned regression tree of depth three is shown in
Figure~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
valley (from east to west, nodes 4 and 5). Secondly, downslope winds blowing
across the Alpine crest along the Wipp Valley towards Innsbruck (from
south-east to north-west, nodes 7 and 8). Thirdly, down-valley winds blowing
in the direction of the valley mouth (from west to east, nodes 12, 14, and 15).
The final node 11 with a concentration parameter $\kappa$ of only $0.84$
captures the cases with rather low wind speeds which typically don't show a
clear wind direction or distinct wind regime. In general, winds typically blow
from areas of high pressure to areas of low pressure where strong pressure
gradients yield high wind speeds and a rather well-defined wind direction.
Thus, the splits in the lagged wind direction perform a clustering into the
different wind regimes and the additional splits in pressure gradients or wind
speed divide the predictions according to the expected consistency of the wind
direction. This can be seen in the final nodes with strong pressure gradients
or high wind speeds where most training cases are classified correctly and
therefore the concentration parameter as an indicator of the certainty of the
forecast is rather large.

\section{Discussion / Summary and Outlook}
This study shows that applying distributional trees on circular data provides easily 
interpretable models clearly illustrating the impact of given covariates and their possible 
interactions on the distribution of the dependent variable. Moreover, covariates are selected
automatically such that no prespecification of the model is required which can be a challenging 
task, particularly for a meteorlogical quantity such as wind direction with high variation and 
many possible influencing factors.

Based on the results it can be assumed that including covariates such as lagged observations of 
wind direction as linear regressors provides valuable information. 
However, in order to properly account not only for the circular properties of the response but 
also of regressors, additional adaptions of distributional trees are required. In particular, 
if the selected split variable is a circular variable, e.g. wind direction, this demands for finding
two split point defining a ``split interval'' instead of a single split point.
This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
the linearly employed regressor \textit{dd} (lagged wind direction) are required in order to 
cluster the training data in corresponding wind directions.

@@

Another possible extension of the method as presented is to build an ensemble of 
distributional trees to regularize and stabilize the model 
(see, e.g., distributional forests, Schlosser et. al, 2018). 
%In that way, it should not have a 
%significant impact on the results whether circular covariates are included as linear regressors
%yielding a larger number of splits or as circular regressors with ``intervall splits''.
